# Housing and Health

> Shelter is one of the three fundamental needs of human existence. No housing program can be sound unless the shelter it provides is healthful.
>
> C.-E. A. Winslow, Dr.P.H., *Chairman*,\
> Allan A. Twichell, *Technical Secretary*,\
> *Committee on the Hygiene of Housing (Mar., 1938)*

## History

The link between health and housing is long and well established.

On January 20, 1852, Dr. John H. Griscom stood before an audience and spoke on the importance of clean air, a topic he believed "was not so much studied as it ought to be." The lecture was summarized on the front page of the New-York Daily Times the next day, concluding with Dr. Griscom's statement that "ventilation in houses should be attended to as one of the best means of preserving health." [@lectures1852]

The idea must have caught on---two and a half years later, an article describing the poor conditions of a building in Buffalo declared, "we intended to speak on the ventilation, but we are sick of the subject" [@thebuff1854]. The home in question had recently been investigated by the local Board of Health based on rumors of high mortality among residents. The authors, after listing the horrid living conditions, concluded on a positive note that they were thankful to "have a Board of Health which we believe will do its duty...to reform this great abuse."

Over the next half century, large population influx into urban areas led to widespread development of tenement housing. [@soifer2014 7] These homes offered little in the way of space, sanitation, light, or fresh air, a fact made well-known by Jacob Riis's 1890 book *How the Other Half Lives*. The conditions documented by Riis led New York City to adopt the first housing code in the United States.

Housing reformers continued to demand improved living conditions into the 20th century. Catherine Bauer's landmark 1934 book *Modern Housing* declared that for a home to be "modern", it must be allow for cross-ventilation and sunlight; adequate privacy, space, and sanitary facilities; adjacent play space for children; and "finally it will be available at a price which citizens of average income or less can afford." The home should be in a "complete neighborhood" and "built for efficient use over a period of years; therefore, it is not designed primarily for quick profits" [@wurster2020].

Bauer was later a primary author on the National Housing Act of 1937, which established public housing authorities and declared it U.S. policy to "remedy the unsafe and unsanitary housing conditions and the acute shortage of decent, safe, and sanitary dwellings for families of low-income, in rural or urban communities, that are injurious to the health, safety, and morals of the citizens of the nation" [@mccarty2019].

The following year, the *Basic Principles of Healthful Housing* was published in the American Journal of Public Health, outlining 30 principles "believed to be fundamental minima required for the promotion of physical, mental, and social health," whether the housing be high or low cost, urban or rural [@apha1938]. The fundamental needs were originally divided into four categories: physiological needs (e.g. protection from elements), psychological needs (e.g. adequate privacy), protection against contagion (e.g. safe water supply), and protection against accidents (e.g. fire prevention).

## Healthy Homes

Six decades after *Healthful Housing* was published, the manual found renewed relevance in 1999. Recognizing that "health, home construction, and home maintenance are inseparable because of their overlapping goals," HUD and CDC joined together to launch the Healthy Homes Initiative [@cdc2006 31, 34]. Many of the basic principles originally laid out in the 1938 manual were used to create the Healthy Home Rating System (HHRS)

The below table lists all 29 hazards of the modern HHRS, along with a short description and which populations are most vulnerable to the specific hazard (adopted from @HHRSlst ).

```{r HHRSdata, include=FALSE}
library(tidyverse)
library(DT)
HHRS <- read_csv("_data/HH-hazards.csv")
```

```{r HHRStbl, echo=FALSE}
datatable(HHRS,
          rownames = FALSE, 
          caption = "The 29 Healthy Home Hazards",
          options = list(
            pageLength = 5,
            lengthMenu = c(5, 10, 15, 20, 25, 29)
          ))
```

<br>

The HHRS provides a standardized way to identify and rate the risk of home health hazards. Once a hazard is identified, it is further rated according to severity and the effect it is having, or could have, on the occupants. The greater the risk or more serious the outcome, the higher the overall score. The system provides a way to compare risks associated with different types of hazards.

According to @HHRSover, this system lets local housing and health departments to know which hazards are most serious to the occupants, allowing them to prioritize funding. It also lets local policy makers identify which areas of the community are in greatest need and what health impacts those communities are facing.

## Specific Health Concerns

The field of public health has dominated academic literature related to the impact of housing quality on residents. The more time a person spends indoors, the more exposed they are to biological, chemical, and physical agents that can affect their health and safety [@cdc2006 63]. Research and findings in the field can lead to national legislation to protect the public health. For instance, the effects of lead-based paint were found to be so severe that all lead-based paint was banned for residential use in 1978 [@cdc2020].

In recent decades, extensive research has been conducted on the relationship between asthma and housing quality. Asthma, a respiratory disease which causes episodes of wheezing, breathlessness, chest tightness, and coughing, has been linked to exposure to mold or dampness and pests including dust mites, cockroaches, and rodents [@cdc2006 63].

Reviewing the existing literature, @mendell2011 found dampness inside homes was linked to a wide range of respiratory or allergic health effects, including asthma. The authors concluded that prevention and remediation to reduce indoor dampness is urgently needed in a large proportion of the building stock, and such measures could significantly reduce the prevalence of respiratory and allergic disease. @rauh2002 found that cockroach allergen levels are related to the degree of household disrepair, and similarly suggested that public health interventions in housing could reduce allergen exposure.

## Maintenance in the Private Rental Market

Health risks caused by larger structural issues can be difficult to implement, particularly for low-income households that rent, as tenants have little or no power to repair such problems. A study of low-income parents of children with asthma found landlords were directly involved in keeping homes in poor condition, even when asked by the tenant to fix the property, and a cycle of fear, poverty, and lack of power compounded to make tenants hesitant to report problems [@grineski2010]. Relocating families can reduce health risks, but higher rents associated with safer housing can make it financially difficult or impossible for some families to move [@mclaine2006].

Current regulation of private rental housing quality assumes that tenants will take action to report substandard housing, yet this is often not the case. To understand the disconnect between the law's expectations and reality, @chisholm2018 collected existing qualitative literature to explore power dynamics in the landlord-tenant relationship. The research showed that, for the most part, tenants who "reported housing quality problems found it a stressful experience, with repairs taking a long time to be carried out, or not at all." This experience often impacted future behavior, causing tenants to avoid reporting problems because they did not think it would be effective.

In some cases, tenants would rather move out than work towards a resolution, leaving the issue open for future tenants. In other instances, tenants with low incomes were aware of the lack of alternative housing, causing them to remain silent. Fear of eviction prevented tenants from reporting problems, and this fear was not unfounded; in three of the 15 studies, tenants who reported housing problems were evicted or forced to move. Though laws against retaliatory action might exist, no-cause evictions allow landlords to still remove the tenant from the unit. @chisholm2018 concludes that tenants do not report housing quality problems because the regulation that relies on their reporting fails to protect many tenants.

The authors propose a wide range of solutions to improve tenant reporting. Ending no-cause evictions and expanding legal aid can encourage tenants who may be scared to speak up. Increasing the supply of social or public housing can expand housing options to households most vulnerable to changes in rent. The authors also suggest authorities supervise a web-based reporting system for repair requests, establishing clear records, automated reminders to the landlords, and an opportunity for authorities to step in for unduly slow repairs.
