# About Memphis

## Housing

The federal government began insuring home loans in the 1930s, and the *Underwriting Manual* was created to outline the rules on how to rate mortgage risk [@rothstein2017]. These manuals contained explicitly racist language, calling for lower ratings in neighborhoods with "inharmonious racial groups".

Memphis was no exception; below is a redlining map of Memphis from the 1930s.

![A color-coded map of Memphis created in the 1930s used by the federal government to rate mortgage risk. Original source: National Archives and Records Administration, accessed via [@bradley2019].](_img/memphis_redliningmap.jpg "Memphis Redlining Map")

Referencing the above map, I created a digitized version using [geojson.io](https://geojson.io/).[^memhsg-1]

[^memhsg-1]: I originally made this digitized map for a class in spring 2021. Today (July 30, 2021) I discovered the University of Richmond's [Mapping Inequality](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58) website, with a plethora of information concerning redlining. HOLC maps are available for public download as original scans and digitized versions (geoJSON or shapefiles) for over 100 cities in the U.S..

```{r libGeo, include=FALSE}
library(rgdal)
library(leaflet)
```

```{r redlineData, include=FALSE}
redData <- "_data/redlining.geojson"
redPoly <- readOGR(redData, 
                    "redlining", 
                    require_geomType = "wkbPolygon",
                    encoding = "UTF-8")
```

```{r redlineLeaflet, echo=FALSE}
leaflet(redPoly) %>% 
  addTiles() %>% 
  addPolygons(stroke = FALSE,
              fillColor = ~ fill,
              fillOpacity = .4,
              label = ~ paste0(Legend, " ", Rating))
```

Overall, 69% of the city was deemed "hazardous" (red) or "definitely declining" (yellow), notably containing all of Downtown and the majority of North and South Memphis. Only 7% of the city received an A rating ("best", green), including the Central Gardens neighborhood, areas adjacent to Overton Park, Chickasaw Gardens, and East Memphis neighborhoods adjacent to golf courses. The remaining 24% of the city ("still desirable", blue) contains other Midtown neighborhoods and follows the Poplar Avenue corridor to the east. Railroads and large streets were typically the dividing line between positive and negative ratings.[^memhsg-2]

[^memhsg-2]: Percentages sourced from [@nelson].

## Health

In 2015, Memphis was named the asthma capital of America by the Asthma and Allergy Foundation of America [@aafa2015], based on prevalence, risk factors, and medical factors. Recent efforts to reduce the prevalence and risk of asthma have been successful---Memphis's rank has since decreased to 41st in 2021 [@aafa2021].[^memhsg-3] However, Memphis still ranks high in asthma-related mortality, placing 7th of 100 cities.

[^memhsg-3]: The ranking method changed between 2015 and 2021, with fewer factors involved in a city's overall ranking in 2021.

A recent study comparing pediatric asthma data with property quality information found that asthma prevalence is disproportionately distributed throughout Memphis [@shin2018]. Neighborhood blight and inequality were closely associated with childhood asthma and other health problems, leading authors to conclude that effective health interventions should consider involving housing policies and procedures.
