# Methodology

Housing maintenance is a universal need. This paper uses the Memphis area (including the City of Memphis, Shelby County, and the Memphis TN-AR-MS MSA) to understand housing maintenance in the private rental market. Information has been divided into three sections: a qualitative housing analysis, a quantitative analysis of available Census data, and an overview of Memphis code enforcement data.

The qualitative section begins by outlining the history of mortgage discrimination in the city, which has denied some Memphians homeownership, leading to forced participation in the rental market. The next portion explores some existing studies and regulation between housing and health in the city. This leads into an overview of the findings of two extensive reports on Memphis code enforcement conducted over the past twenty years.

A quantitative analysis of Census data uses information from 2010-2019 American Community Surveys (ACS) and the 2019 American Housing Survey (AHS) to understand certain dynamics in Memphis's housing market. Particular attention is given to tenure, as the city's households recently switched from being majority owner-occupied to majority renter-occupied. Data on building age adds credence to the argument that housing maintenance needs attention, particularly with analysis showing the conversion of old, formerly owner-occupied units to rentals. The section concludes with AHS data showing that renter-occupied homes experience a consistently higher rate of housing problems and deficiencies than owner-occupied homes.

The final section centers around a dataset of service requests to the City of Memphis from 2016-present, including requests to code enforcement. The original intent of using this data was to study the extent of housing maintenance requests in the private rental market and how problems were addressed. However, a larger problem emerged of the dataset's general disorganization. With no known existing methodology for how to parse this data, this paper serves to introduce other housing researchers to specific problems with the dataset, with opportunity for analysis and collaboration moving forward.

# About Memphis

Memphis and Shelby County were the recent focus of a year-long study of evictions by Legal Services Corporation, a publicly-funded provider of civil legal aid to low-income Americans [@ahmed2021]. The area was selected not because it is an outlier---"rather because it is typical of many U.S. counties," with a population concentrated in a major urban center and a "not exorbitant" cost of housing.

Multiple factors make Memphis ideal for monitoring maintenance in the private rental market. The housing stock in the central city is old; the median year built for owner and renter-occupied homes is before 1978. Not only will these homes require additional maintenance, but homes built before this time likely used lead-based paint. Unless the property has been remediated, the paint can create lead dust in the home, a particular hazard for young children [@cdc2020].

Mortgage lending discrimination has disproportionately denied non-white residents the ability to purchase homes or build wealth---not just as a part of redlining practices, but also in recent years before and after the Great Recession. The fallout of the recession caused high foreclosure rates, transitioning Memphis from majority homeowner to majority renter city. In the following decade, Memphis tenants saw rents outpace incomes, leading to increased cost burdenship.

Existing in a decidedly pro-landlord state has limited the protections of Memphis tenants. This influences the decisions of renters, who must weigh decisions like reporting their landlord to code enforcement at the expense of retaliation/eviction. This can lead tenants to under-reporting problems, which is something that should be looked for and measured by housing researchers.

Recent changes to Memphis's comprehensive plan also make the city an interesting case study. In other places across America, stringent zoning laws have limited the supply of housing, and discourse over affordable housing policies has been narrowly focused on changing or repealing these zoning laws. In 2019, Memphis adopted the comprehensive plan known as "Memphis 3.0," which relaxed zoning codes to encouraged mixed use development and increased density in the city [@memphis30]. This change allows researchers to study the effects of relaxed zoning and consider other factors influencing affordable housing that are less discussed.

Lastly, code enforcement data is easily accessed from [Memphis Data Hub](https://data.memphistn.gov/dataset/Service-Requests-since-2016/hmd4-ddta), where it is updated daily.

## Mortgage Discrimination

### Historical

The federal government first became involved with mortgage loans in the 1930s through the Home Owner's Loan Corporation (HOLC) and Federal Housing Administration (FHA). HOLC refinanced loans at risk of default following the depression, and commissioned color-coded maps rating the mortgage risk of neighborhoods in every major metro area across America. Risk was in part determined by a neighborhood's racial composition. Black neighborhoods were colored red for high-risk, even if the neighborhood was solidly middle class and single-family homes [@rothstein2017 64].

The FHA made first-time homeownership more accessible to middle-class renters by instituting the idea of 20-year mortgages, rather than the prohibitively expensive 5- to 7-year ones common at the time. If borrowers defaulted, the FHA insured lenders for full repayment [@mccarty2019; @rothstein2017 63-64]. To make sure a loan had a low-risk of default, the FHA often hired a local real estate agent to conduct a property appraisal. The FHA created the *Underwriting Manual* to instruct agents on how to rate mortgage risk [@rothstein2017 65].

These manuals prioritized newly built suburban neighborhoods and discouraged banks from making any loans in urban areas. They contained explicitly racist language, calling for lower ratings in neighborhoods with "inharmonious racial groups" [@federalhousingadministration1938 §§ 935, 937, 951, 982]. This language was dropped in 1947, though the manual continued to value properties based on "compatibility among the neighborhood occupants" [@rothstein2017 66, citing the 1952 *Manual*].

Color-coded maps were created for cities across the U.S. to illustrate mortgage ratings. Neighborhoods deemed the riskiest were shaded red, leading to the term "redlining." Memphis was no exception; the following is a mortgage rating map from the 1930s.

![A color-coded map of Memphis created in the 1930s used by the federal government to rate mortgage risk. Original source: National Archives and Records Administration, accessed via [@bradley2019].](_img/memphis_redliningmap.jpg "Memphis Redlining Map")

A digitized version of the above map is available below.[^memhsg-1]

[^memhsg-1]: I originally made this map using [geojson.io](https://geojson.io/) for a class in spring 2021. I have since discovered the University of Richmond's [Mapping Inequality](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58) website, which offers original scans and digitized versions (geoJSON or shapefiles) of redlining maps available for download for over 100 U.S. cities.

```{r redlineData, include=FALSE}
library(rgdal)
library(leaflet)
redData <- "_data/redlining.geojson"
redPoly <- readOGR(redData, 
                    "redlining", 
                    require_geomType = "wkbPolygon",
                    encoding = "UTF-8")
```

```{r redlineLeaflet, echo=FALSE}
leaflet(redPoly) %>% 
  addTiles() %>% 
  addPolygons(stroke = FALSE,
              fillColor = ~ fill,
              fillOpacity = .4,
              label = ~ paste0(Legend, " ", Rating))
```

<br></br>

Overall, 69%[^memhsg-2] of the city was deemed "hazardous" (red) or "definitely declining" (yellow), notably including all of Downtown and the majority of North and South Memphis. The federal government deemed these mortgages risky, and either denied the loan or charging higher rates to account for the "risk".

[^memhsg-2]: Percentages sourced from [@nelson].

Only 7% of the city received an A rating ("best", green), including the Central Gardens neighborhood, areas adjacent to Overton Park, Chickasaw Gardens, and East Memphis neighborhoods adjacent to golf courses. These neighborhoods have notably remained occupied and maintained over the decades.

The remaining 24% of the city ("still desirable", blue) contains other Midtown neighborhoods and follows the Poplar Avenue corridor to the east, drawing residents toward the suburbs.

### Modern Discrimination

In recent years, at least three major lawsuits have been settled concerning discriminatory lending in Memphis [@bradley2019; @uhlmann2020]. Excessive foreclosures following the 2008 subprime mortgage crisis led Memphis and Shelby County to file suit against Wells Fargo for targeting Black neighborhoods for predatory lending practices, which the bank settled for \$425 million in 2012 [@rothacker2012].

Following the Great Recession, banks became selective of whom they targeted and approved for mortgages, often at the expensive of non-white residents. In 2016, BancorpSouth and First Tennessee Bank settled separate suits for \$10.6 million and \$1.9 million, respectively, against accusations of disproportionately denying loan applications to non-white borrowers and failing to build branches in non-white neighborhoods [@lane2016a; @lane2016].

## Health & Healthy Homes

In 2015, Memphis was named the asthma capital of America by the Asthma and Allergy Foundation of America [@aafa2015], based on prevalence, risk factors, and medical factors. Recent efforts to reduce the prevalence and risk of asthma have been successful---Memphis's rank has since decreased to 41st in 2021 [@aafa2021].[^memhsg-3] However, Memphis still ranks high in asthma-related mortality, placing 7th of 100 cities. A recent study comparing pediatric asthma data with property quality information found that asthma prevalence is disproportionately distributed throughout Memphis [@shin2018]. Neighborhood blight and inequality were closely associated with childhood asthma and other health problems.

[^memhsg-3]: The ranking method changed between 2015 and 2021, with fewer factors involved in a city's overall ranking in 2021.

In the Memphis area, the Healthy Homes Rating System has been used by inspectors with the Shelby County Department of Housing. The agency's Lead Hazard Reduction Program remediates homes with elevated lead levels. Funding is provided by HUD, and prior to beginning repairs, HUD requires the home be inspected for health hazards. Any funds leftover from lead remediation can be used towards improving Healthy Home violations.

## Code Enforcement

Over the past twenty years, there have been at least two expansive reports offering critiques and criticisms of Memphis code enforcement.

The first report, released in April 2001, was distributed by the Memphis Shelby Crime Commission and written by Phyllis Betts, then a professor of sociology at University of Memphis [@betts2001]. Input was provided by two other University of Memphis faculty: Betts's husband, the late Richard Janikowski, former chair of the Department of Criminology and Criminal Justice and 'father' of Memphis's Blue CRUSH policing [@poe2021]; and Susan Roakes, former professor of City and Regional Planning.

The report was built on the "broken windows" criminology theory: in deteriorated or declining neighborhoods, physical neglect of "problem properties" attracts and aggravates criminal activity [@betts2001]. Researchers interviewed inspectors and individuals involved with environmental court, used code enforcement data from 1992-1999, accessed the Shelby County Tax Assessor's database, and conducted visual surveys for selected properties as part of a case study of the Binghamton neighborhood.

The second report, published in 2018, was produced by The Urban Institute [@stacy2018]. While the 2001 report focused on property neglect and *safety*, this report centered on the link between the physical condition of homes and neighborhoods and *public health*. Planning for the report began shortly after the election of Mayor Jim Strickland in 2016, whose campaign had emphasized blight control [@baker2018]. Researchers collaborated with and interviewed members of the newly formed [Blight Elimination Steering Team](http://memphisfightsblight.com/blight-elimination-steering-team/), a coordinated effort between public, private, and nonprofit agencies to address blight, and the [Green and Healthy Homes Initiative](https://www.greenandhealthyhomes.org/location/memphis-tn/), a partnership formed between local health care providers and housing and community development organizations to address public health issues related to housing quality.

The Urban report accessed and analyzed data through the [Memphis Property Hub](http://memphisfightsblight.com/programs/mph/), which offers parcel-level property information, particularly related to blighted, vacant, and foreclosed homes. Code enforcement data from 1999-2017 was used, effectively picking up where the last report ended. Additional data related to public health, safety, and courts were provided by the Shelby County Healthy Department, the Center for Applied Earth Science and Engineering Research department at the University of Memphis, and the Shelby County Environmental Court, respectively.

### Findings and Recommendations

Despite nearly twenty years difference, the reports contain similar finding and recommendations. Notably, each report recommended fundamentally changing Memphis code enforcement from a reactive process (responding to citizen reports) to a proactive, strategic system. This was largely due to the large volume of cases processed related to vehicle violations or weeds, rather than more serious violations.

Both reports found the vast majority of code enforcement reports were related to nonstructural issues. @betts2001 reported three out of four properties were cited for non-structural problems such as weeds, junk, and inoperable or abandoned vehicles. It appears this percentage has actually increased overtime; @stacy2018 stated 81% of all requests concerned the yard or property, rather than the house.

@stacy2018 found service requests were concentrated in areas with single-family homes, and though about half of all requests were for multifamily units, most of these (70 percent) were for duplexes rather than apartment complexes. @betts2001 also found that large multifamily units flew under the radar; at least half of multifamily properties in Binghamton had violations in multiple units, yet code enforcement recorded only eight violations out of 1200 apartment units.

When the report's findings differed, it was usually due to increased workload. The number of service requests has appeared to increase overtime. In 1998, code enforcement received about 10,000 complaints [@betts2001 93], while the Urban Institute reported 53,226 in 2016 [@stacy2018 51]. However, this number does not match a dataset from the City of Memphis (discussed later in this paper) which reported 30,361 requests in 2016, not excluding duplicates. Nevertheless, it does appear that more requests are being made.

Newer cases also appear significantly more likely to be taken to court. In 1998, a mere 71 cases from code enforcement were heard by the environmental court [@betts2001 49]. In November 2017 alone, 1,095 cases went to environmental court, though the Urban report does not specifically state if all cases were from code enforcement [@stacy2018 51]. The majority of new cases were for single-family homes (64%), and most were vacant.

@stacy2018 suggests that code enforcement prioritize health-related violations and interior health and safety issues, cemented through the agency's policies and procedures manual. The authors recommend proactive inspections of problem multifamily properties and under-served neighborhoods, and identification of under-reported violations. Multiple technology improvements are recommended, including automatic prioritization of health-related violations and synchronizing of data systems with the Shelby County Department of Health and the environmental court, among others. Also acknowledged is the need to update current landlord/tenant laws to protect tenants from homelessness.
