# About Memphis

## Housing

The federal government began insuring home loans in the 1930s, and the *Underwriting Manual* was created to outline the rules on how to rate mortgage risk [@rothstein2017 65]. These manuals prioritized newly built suburban neighborhoods, discouraging banks from making any loans in urban areas. They contained explicitly racist language, calling for lower ratings in neighborhoods with "inharmonious racial groups" [@federalhousingadministration1938 §§ 935, 937, 951, 982]. This language was dropped in 1947, though the manual continued to value properties based on "compatibility among the neighborhood occupants" [@rothstein2017 66, citing the 1952 *Manual*]. Color-coded maps were created for cities across the U.S. to illustrate mortgage ratings. Neighborhoods deemed the riskiest were shaded red, leading to the term "redlining."

Memphis was no exception; below is a mortgage rating map from the 1930s.

![A color-coded map of Memphis created in the 1930s used by the federal government to rate mortgage risk. Original source: National Archives and Records Administration, accessed via [@bradley2019].](_img/memphis_redliningmap.jpg "Memphis Redlining Map")

Below is a digitized version of the above map.[^memhsg-1]

[^memhsg-1]: I originally made this map using [geojson.io](https://geojson.io/) for a class in spring 2021. I have since discovered the University of Richmond's [Mapping Inequality](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58) website, which offers original scans and digitized versions (geoJSON or shapefiles) of redlining maps available for download for over 100 U.S. cities.

```{r libGeo, include=FALSE}
library(rgdal)
library(leaflet)
```

```{r redlineData, include=FALSE}
redData <- "_data/redlining.geojson"
redPoly <- readOGR(redData, 
                    "redlining", 
                    require_geomType = "wkbPolygon",
                    encoding = "UTF-8")
```

```{r redlineLeaflet, echo=FALSE}
leaflet(redPoly) %>% 
  addTiles() %>% 
  addPolygons(stroke = FALSE,
              fillColor = ~ fill,
              fillOpacity = .4,
              label = ~ paste0(Legend, " ", Rating))
```

<br>

Overall, 69% of the city was deemed "hazardous" (red) or "definitely declining" (yellow), notably containing all of Downtown and the majority of North and South Memphis. Only 7% of the city received an A rating ("best", green), including the Central Gardens neighborhood, areas adjacent to Overton Park, Chickasaw Gardens, and East Memphis neighborhoods adjacent to golf courses. The remaining 24% of the city ("still desirable", blue) contains other Midtown neighborhoods and follows the Poplar Avenue corridor to the east.[^memhsg-2] Railroads and large streets were typically the dividing line between positive and negative ratings.

[^memhsg-2]: Percentages sourced from [@nelson].

In recent years, at least three major lawsuits have been settled concerning discriminatory lending in Memphis [@bradley2019; @uhlmann2020]. Excessive foreclosures following the 2008 subprime mortgage crisis led Memphis and Shelby County to file suit against Wells Fargo for targeting Black neighborhoods for predatory lending practices, which the bank settled for \$425 million in 2012 [@rothacker2012].

Following the Great Recession, banks became selective of whom they targeted and approved for mortgages, often at the expensive of non-white residents. In 2016, BancorpSouth and First Tennessee Bank settled separate suits for \$10.6 million and \$1.9 million, respectively, against accusations of disproportionately denying loan applications to non-white borrowers and failing to build branches in non-white neighborhoods [@lane2016a and @lane2016].

## Health

In 2015, Memphis was named the asthma capital of America by the Asthma and Allergy Foundation of America [@aafa2015], based on prevalence, risk factors, and medical factors. Recent efforts to reduce the prevalence and risk of asthma have been successful---Memphis's rank has since decreased to 41st in 2021 [@aafa2021].[^memhsg-3] However, Memphis still ranks high in asthma-related mortality, placing 7th of 100 cities.

[^memhsg-3]: The ranking method changed between 2015 and 2021, with fewer factors involved in a city's overall ranking in 2021.

A recent study comparing pediatric asthma data with property quality information found that asthma prevalence is disproportionately distributed throughout Memphis [@shin2018]. Neighborhood blight and inequality were closely associated with childhood asthma and other health problems, leading authors to conclude that effective health interventions should consider involving housing policies and procedures.
